---
title: "Practical Machine Learning Coursera Project"
author: "Mrinalini Prakhya"
date: "January 26, 2017"
output: html_document
---
  
  ##Background
  
  Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Data

The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har]. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

##Goal

The goal of your project is to predict the manner in which they did the exercise. This is the classe variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.
Submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-). You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

##Approach:

###Reproducibility

An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used.
Different packages were downloaded and installed, such as caret and randomForest and rpart. They are installed in order to reproduce the results.

Our outcome variable is classe, a factor variable. For this data set, participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:  
  1. exactly according to the specification (Class A)  
2. throwing the elbows to the front (Class B) 
3. lifting the dumbbell only halfway (Class C)  
4. lowering the dumbbell only halfway (Class D) 
5. throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.

Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using Decision tree and Random forest algorithms. The model with the highest accuracy will be chosen as our final model.

###Cross-validation

Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: TrainTrainingSet data (75% of the original Training data set) and TestTrainingSet data (25%). Our models will be fitted on the TrainTrainingSet data set, and tested on the TestTrainingSet data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

###Expected out-of-sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the TestTrainingSet data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

Reasons for my choices

Our outcome variable classe is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. We have a large sample size with N= 19622 in the Training data set. This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation. Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification [2]. Feature selection is inherent, so it is not so necessary at the data preparation phase. Thus, there is no feature selection section in this report.


Install packages and load the required libraries

install.packages("caret"); 
install.packages("randomForest"); 
install.packages("rpart"); 
```{r}
#load libraries
library(lattice) 
library(ggplot2)
library(munsell)
library(caret) 
library(randomForest) 
library(rpart) 
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(data.table)
library(survival)
library(splines)
library(RCurl)
library(plyr)
library(dplyr)
library(lubridate)



set.seed(1234)
```

##Loading data sets and preliminary cleaning

First we want to load the data sets into R and make sure that missing values are coded correctly.
Unnecessary variables will be deleted.
Results will be hidden from the report for clarity and space considerations.

Saving both data sets into the working directory
Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
We notice that both data sets contain columns with all missing values - they will be deleted.  


##Data loading and clean up

```{r }
if(!file.exists("trainingdata")) {
  dir.create("trainingdata")}
fileurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileurl,destfile = "C:/Users/Vamsi/Desktop/Coursera/Mrinalini/Practical-MachineLearning/Trainingdata.csv")
if(!file.exists("testingdata")) {
  dir.create("testingdata")}
fileurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileurl,destfile = "C:/Users/Vamsi/Desktop/Coursera/Mrinalini/Practical-MachineLearning/Testingdata.csv")

# Loading the training data set into R and replacing all missing with "NA"
trainingset <- read.csv("C:/Users/Vamsi/Desktop/Coursera/Mrinalini/Practical-MachineLearning/Trainingdata.csv", na.strings=c("NA","#DIV/0!",  col_types = "_nni"))

# Loading the testing data set  
testingset <- read.csv("C:/Users/Vamsi/Desktop/Coursera/Mrinalini/Practical-MachineLearning/Testingdata.csv", na.strings=c("NA","#DIV/0!",  col_types = "_nni"))
```


### Perform exploratory analysis -

dim(trainingset);dim(testingset); summary(trainingset); summary(testingset); str(trainingset); str(testingset); head(trainingset); head(testingset);               

### Delete columns with all missing values

```{r}
trainingset <- trainingset[,colSums(is.na(trainingset)) == 0]
testingset <- testingset[,colSums(is.na(testingset)) == 0]
```

###Delete variables that are not relevant to the current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7) 
```{r}
trainingset <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]
```

### New datasets:
```{r}
dim(trainingset)
dim(testingset)
head(trainingset)
head(testingset)
```

###Partition the data so that 75% of the training dataset into training and the rest 25% to testing

The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.

```{r}
subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)
```

The variable classe contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.
```{r}
plot(subTraining$classe, col="grey", main="Bar Plot of the levels of variable classe in the SubTraining dataset", xlab="classe levels", ylab="Frequency")
```
From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

###Prediction model 1: Decision Tree
```{r}
model1 <- rpart(classe ~ ., data=subTraining, method="class")

prediction1 <- predict(model1, subTesting, type = "class")
```
### Plot the Decision Tree
```{r}
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

###Test results on Subtesting data set:
```{r}
confusionMatrix(prediction1, subTesting$classe)
```

###Prediction model 2: Random Forest
```{r}
model2 <- randomForest(classe ~. , data=subTraining, method="class")
```

### Predicting:
```{r}
prediction2 <- predict(model2, subTesting, type = "class")
```
### Test results on TestTrainingSet data set:
```{r}
confusionMatrix(prediction2, subTesting$classe)
```

###Decision on which Prediction Model to Use:
Random Forest algorithm performed better than Decision Trees. Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to Decision Tree model with 0.739 (95% CI: (0.727, 0.752)). The Random Forests model is choosen. The expected out-of-sample error is  0.005 or 0.5%.

Final outcome
Here is the final outcome based on the Prediction Model 2 (Random Forest) applied against the Testing dataset


###Predict outcome levels on the original Testing data set using Random Forest algorithm
```{r}
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```